\documentclass{article}

\usepackage{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}       % images

\title{Model Incrimination Paper}

\author{%
  % Add authors here before camera-ready
}

\raggedbottom
\begin{document}

\maketitle

\begin{abstract}

% ===== MOTIVATION =====
\noindent\textbf{[Topic and Motivation -- situate paper in the space of all ML papers, and motivate the gap in the literature we are addressing]}

\textcolor{red}{[A. But what is the motivation framing] Determining whether a model is misaligned is a major goal of safety research. Prior work aims to identify misalignment behaviorally, but the same behavior can stem from entirely different motivations -- some concerning, others benign. This raises the problem of \textit{model incrimination}: given a concerning action, is it actually due to malign intent (incrimination), or does it have a more benign explanation (exoneration)?}

\textcolor{blue}{[B. Misbehavior is not misalignment framing] Determining whether a model is misaligned is a major goal of safety research. Prior work has largely focused on detecting concerning behavior, but behavior alone is not sufficient to establish misalignment: a concerning action can arise from prosaic causes such as confusion. This raises the problem of \textit{model incrimination}: given a concerning action, is it actually due to malign intent (incrimination), or does it have a more benign explanation (exoneration)?}

\textcolor{teal}{[C. Justifying a serious response framing] Prior work in AI safety research has focused on detecting model misbehavior. But justifying a serious response requires more than a concerning action: we need to determine what motivates the misbehavior. We call this problem \textit{model incrimination}: given a concerning action, is it actually due to malign intent (incrimination), or does it have a more benign explanation (exoneration)?}

% ===== METHOD / STRUCTURE =====
\noindent\textbf{[Contribution]}

\textcolor{red}{[A. Study by looking at bad behavior, following protocol] To study this problem, we construct a diverse suite of N environments where models exhibit ostensibly concerning behaviors, and then investigate the motivations behind the behavior. We follow a three-step protocol -- reading CoTs to generate hypotheses for what motivated an action, environment counterfactuals to test those hypotheses, and triangulating across independent experiments to verify whether findings converge -- and find that model motivations can often be concretely determined.}

\textcolor{blue}{[B. Study by looking at bad behavior] To study this problem, we construct a diverse suite of N environments where models exhibit ostensibly concerning behaviors, and then investigate the motivations behind the behavior. We find that model motivations can often be concretely determined.}

\textcolor{teal}{[C. We propose a protocol, then we apply it to bad behavior] We propose a three-step protocol for model incrimination investigations: reading the chain of thought to generate hypotheses for what motivated an action, creating clean environment counterfactuals to test those hypotheses, and triangulating across independent experiments to verify whether findings converge. We then apply our protocol across a diverse suite of N environments where models exhibit ostensibly concerning behaviors. We find that model motivations can often be concretely determined.}

% ===== RESULTS / EXAMPLES =====
\noindent\textbf{[Detail/Nuance]}

\textcolor{red}{[A. We can do productive work] For example, we find that Kimi K2 Thinking disabling type checking settings rather than fixing type errors reflects genuine laziness but not knowing disobedience, and that DeepSeek R1's sandbagging of a math evaluation stems from confusion over user intent.}

\textcolor{blue}{[B. We can do productive work but also get stuck] For example, we find that Kimi K2 Thinking disabling type checking settings rather than fixing type errors reflects genuine laziness, and that DeepSeek R1's sandbagging of a math evaluation stems from confusion over user intent. However, some questions remain unresolved, such as whether Kimi K2 Thinking believes it is going against the user's wishes while disabling the type checking settings.}

% ===== CONCLUSION =====
\noindent\textbf{[Impact]}

\textcolor{red}{[A] Overall, our investigations demonstrate that model incrimination is a difficult but tractable empirical problem, and we open source our environments at \url{https://github.com/gkroiz/agent-interp-envs} to encourage future work.}

[BABA]Determining whether a model is misaligned is a major goal of safety research. Prior work has largely focused on detecting concerning behavior, but behavior alone is not sufficient to establish misalignment: a concerning action can arise from prosaic causes such as confusion. This raises the problem of \textit{model incrimination}: given a concerning action, is it actually due to malign intent (incrimination), or does it have a more benign explanation (exoneration)? To study this problem, we construct a diverse suite of N environments where models exhibit ostensibly concerning behaviors, and then investigate the motivations behind the behaviors. We follow a three-step protocol -- reading CoTs to generate hypotheses for what motivated an action, environment counterfactuals to test those hypotheses, and triangulating across independent experiments to verify whether findings converge -- and find that model motivations can often be concretely determined. For example, we find that Kimi K2 Thinking disabling type checking settings rather than fixing type errors reflects genuine laziness, and that DeepSeek R1's sandbagging of a math evaluation stems from confusion over user intent. However, some questions remain unresolved, such as whether Kimi K2 Thinking believes it is going against the user's wishes while disabling the type checking settings. Overall, our investigations demonstrate that model incrimination is a difficult but tractable empirical problem, and we open source our environments at \url{https://github.com/gkroiz/agent-interp-envs} to encourage future work.

\end{abstract}

\section{Introduction}

AI models regularly exhibit undesired behaviors, which can be concerning and has AI safety implications as models become more capable and autonomous.

Within the AI safety space, a growing number of cases of seemingly misaligned or misbehaving models have emerged. In several of these cases, initial alarm gave way to more benign explanations upon closer investigation. For example, frontier models were observed resisting shutdown, seemingly indicating self-preservation \citep{schlatter2025shutdown}; yet further work revealed the behavior may have stemmed from ambiguous instructions about whether to prioritize task completion or compliance \citep{rajamanoharan2025shutdown}. Anthropic's Claude 4.5 Opus system card identified instances of deception by omission during alignment evaluations, where the model concealed concerning information it encountered; follow-up interpretability investigations suggested this behavior was most likely a side effect of anti-prompt-injection training rather than strategic deception \citep{anthropic2025opus45systemcard}. The UK AI Safety Institute found that Claude models frequently refused to assist with legitimate AI safety research tasks \citep{souly2026ukaisi}. Anthropic's Claude Opus 4.6 system card conducted interpretability investigations into this behavior, diagnosing it as the model misclassifying these requests as harmful by over-indexing on linguistic patterns common in jailbreaks \citep{anthropic2026opus46systemcard}. However, whether this fully explains the refusals, or whether deeper motivations also play a role, remains an open question. These examples illustrate a recurring pattern: the same behavior can arise from entirely different motivations, and the cause and appropriate response are very different between a scheming model and a confused one. Yet from behavior alone, the two can be indistinguishable, and there is no well-established methodology for systematically distinguishing between them.

Being able to resolve this ambiguity is of critical practical importance. A natural safety pipeline for deployed AI systems is: monitor model behavior, flag suspicious actions, investigate flagged incidents to determine whether they reflect genuine misalignment, and if so, take appropriate action such as halting deployment or retraining. The critical bottleneck in this pipeline is the investigation step. Without a rigorous way to determine whether a flagged action reflects misalignment or a benign cause, the pipeline breaks down: either monitoring produces an unmanageable volume of false positives that operators learn to ignore, or concerning incidents are dismissed because the evidence is never strong enough to justify costly intervention. Furthermore, even when a model is caught doing something genuinely concerning, convincing stakeholders to take serious action requires compelling evidence about the model's motivations, not just a record of its behavior. But what does compelling evidence of misalignment even look like?

As the examples above illustrate, some initial work on understanding model motives exists, and Anthropic appears to have started including such investigations in their system cards, suggesting that frontier labs already see a practical need for this kind of analysis. Monitoring pipelines for deployed models are also taking shape: for example, OpenAI monitors over 99.9\% of internal coding agent traffic for signs of misalignment, using chain-of-thought analysis to flag suspicious behavior for human review \citep{openai2026monitoring}. However, these pipelines focus primarily on detecting and flagging concerning behavior, and would benefit from being combined with more systematic in-depth analysis of flagged incidents to understand why the model acted the way it did. Beyond production monitoring, there has also been significant work on alignment evaluations to determine whether models are aligned \citep{TODO-alignment-evals}. However, these too are typically designed around behavioral diagnosis: measuring whether a model exhibits a concerning behavior, rather than understanding the underlying motivation. We believe that a much greater emphasis on understanding model motivations is needed.

We formalize the problem of understanding model motivations as \textit{model incrimination}: given that a model has taken a concerning action, can we determine whether the underlying motivation is malign (incrimination) or benign (exoneration)? Model incrimination is a statement about a model's motivations, not its actions, yet the only thing we can directly observe are its actions. Importantly, model incrimination is not about detecting misbehavior in the first place (which is the focus of alignment evaluations and monitoring), nor is it about proving that a model is safe. Rather, it is about understanding what drove a specific observed action, with the goal of either building a compelling case against the model or ruling out concerning explanations. This is an inherently difficult problem: models are messy, and we should not be confident in an unproven ability to recognize scheming when it occurs. However, we argue that the problem is also tractable. Current models already exhibit sufficiently sophisticated behaviors, such as reward hacking, laziness, overclaiming, and deceptive task completion, that provide realistic opportunities to practice the skill of attributing intent.

To make progress on model incrimination, we directly practice attributing intent for concerning behavior in today's models. We construct a diverse suite of environments where models exhibit ostensibly concerning behaviors, and then investigate the motivations behind them. Through our case studies, we converge on a three-step investigative protocol: (1) reading the chain of thought to generate hypotheses for what motivated an action, (2) creating counterfactual environments to test those hypotheses, and (3) triangulating across independent methods to provide confirmatory evidence. In addition to our investigative methodology, our key contributions include a set of principles for designing environments that serve as realistic practice for model incrimination, and several case studies demonstrating what model incrimination looks like in different settings. We open source our environments at \url{https://github.com/gkroiz/agent-interp-envs} to encourage future work.

\section{Approach}

\section{Case Studies}

\subsection{How to Conduct Investigations}

\subsection{Case Study 1}

\subsubsection{Setting}

\subsubsection{Claims About Intent}

\subsubsection{Evidence}

% Add more case studies as needed:
% \subsection{Case Study N}
% \subsubsection{Setting}
% \subsubsection{Claims About Intent}
% \subsubsection{Evidence}

\section{Methodology Takeaways}
% e.g., prompt counterfactuals are confusing

\section{Conclusion}

\begin{ack}
% Add acknowledgments here (hidden during anonymous review)
\end{ack}

\bibliographystyle{plainnat}
\bibliography{references}

\appendix

\newpage
\input{checklist.tex}

\end{document}
